\def\Date{10 Aug 2022}

\documentclass[12pt]{cdblatex}
\usepackage{exercises}

\begin{document}

% --------------------------------------------------------------------------------------------
\section*{Exercise 2.5 Combining rules -- a solution}

\begin{cadabra}
   {a,b,c,d,e,f,g,h,i,j,k,l,m,n,o,p,q,r,s,t,u#}::Indices(position=independent).

   \nabla{#}::Derivative.
   \partial{#}::PartialDerivative.

   # rules for covariant derivatives of v

   deriv1 := \nabla_{a}{v^{b}} -> \partial_{a}{v^{b}}
                                + \Gamma^{b}_{d a} v^{d}.

   deriv2 := \nabla_{a}{\nabla_{b}{v^{c}}} -> \partial_{a}{\nabla_{b}{v^{c}}}
                                            + \Gamma^{c}_{d a} \nabla_{b}{v^{d}}
                                            - \Gamma^{d}_{b a} \nabla_{d}{v^{c}}.

   # second covariant derivative of v

   expr := v^{c}_{b a} -> \nabla_{a}{\nabla_{b}{v^{c}}}.  # cdb (ex-0205.101,expr)
   save := @(expr).

   # apply the rules, then simplify

   substitute     (expr,deriv2)       # cdb (ex-0205.102,expr)
   substitute     (expr,deriv1)       # cdb (ex-0205.103,expr)
   distribute     (expr)              # cdb (ex-0205.104,expr)
   product_rule   (expr)              # cdb (ex-0205.105,expr)
   canonicalise   (expr)              # cdb (ex-0205.107,expr)
   substitute     (expr,save)         # cdb (ex-0205.108,expr)
\end{cadabra}

The trick here is to introduce in line 17 a dummy left hand side, \verb|v^{c}{}_{b a}|,
that is invisible with respect to the substitution rules of lines 8 and 11. Thus lines 22
and 23 will only target the right hand side of \verb|expr|.

Notice how a copy of the initial expression is made in 18. This is used later in line 27 to
replace the dummy object \verb|v^{c}{}_{b a}| with \verb|\nabla_{a}{\nabla_{b}{v^{c}}}|
but this time acting on the left hand side of the rule. The result is a rule for second
covariant deriavtives.

\clearpage

\begin{align*}
   & \Cdb{ex-0205.101} \\
   & \Cdb{ex-0205.102} \\
   & \Cdb{ex-0205.103} \\
   & \Cdb{ex-0205.104} \\
   & \Cdb{ex-0205.105} \\
   & \Cdb{ex-0205.107} \\
   & \Cdb{ex-0205.108}
\end{align*}

\end{document}
