\documentclass[12pt]{cdblatex}
\usepackage{exercises}

\begin{document}

% --------------------------------------------------------------------------------------------
\section*{Exercise 4.5 Reformatting complex expressions}

\begin{cadabra}

   {a,b,c,d,e,f,g,h,i,j,k,l,m,n,o,p,q,r,s,t,u,v,w#}::Indices(position=independent).

   \nabla{#}::Derivative.

   def get_term (obj,n):

       x^{a}::Weight(label=xnum).     # assign weights to x^{a}

       foo := @(obj).                 # make a copy of obj
       bah  = Ex("xnum = " + str(n))  # choose a target
       keep_weight (foo,bah)          # extract the target

       return foo

   def reformat (obj,scale):

       {x^{a},A_{a},B_{a},A_{a b},B_{a b},C_{a b},C_{a b c},g^{a b}}::SortOrder.  # choose a sort order

       foo  = Ex(str(scale))          # create a scale factor
       bah := @(foo) @(obj).          # apply the scale factor, clears all fractions

       distribute     (bah)           # only required if (bah) contains brackets
       sort_product   (bah)
       rename_dummies (bah)
       canonicalise   (bah)
       factor_out     (bah,$x^{a?}$)

       ans := @(bah) / @(foo).        # undo previous scaling

       return ans

   # ---------------------------------------------------------------

   # a messy unformatted expression

   expr :=    (1/7) A_{e} x^{e}
           -  (1/3) B_{f} x^{f}
           +  (1/3) A_{a b} x^{a} x^{b}
           +  (1/9) B_{e c} x^{c} x^{e}
           -  (1/5) C_{p c} B_{d q} g^{c d} x^{p} x^{q}
           +  (3/7) A_{a b c} x^{a} x^{b} x^{c}
           -  (1/5) B_{a b} C_{c d e} g^{c d} x^{a} x^{b} x^{e}
           + (7/11) B_{a b} B_{c d} C_{e f g} g^{b c} g^{d f} x^{a} x^{e} x^{g}.  # cdb (ex-0405.100,expr)

   # split the expression into seprate terms

   term1 = get_term (expr,1)        # cdb(term1.101,term1)
   term2 = get_term (expr,2)        # cdb(term2.101,term2)
   term3 = get_term (expr,3)        # cdb(term3.101,term3)

   # reformat terms and tidy fractions

   term1 = reformat (term1, 21)     # cdb(term1.102,term1)
   term2 = reformat (term2, 45)     # cdb(term2.102,term2)
   term3 = reformat (term3,385)     # cdb(term3.102,term3)

   # rebuild the expression

   expr := @(term1) + @(term2) + @(term3).  # cdb (ex-0405.101,expr)

\end{cadabra}

\clearpage

\begin{dgroup*}
   \Dmath*{ g = \Cdb*[\V{15pt}\hskip 3.0cm\hfill]{ex-0405.100}
              = \Cdb*{ex-0405.101} }
\end{dgroup*}

\end{document}
