\documentclass[12pt]{cdblatex}
\usepackage{examples}
\usepackage{fancyhdr}
\usepackage{footer}

\begin{document}

% ============================================================================================
\bgroup
\CdbSetup{action=hide}
\begin{cadabra}
   import cdblib
   checkpoint_file = 'tests/semantic/output/example-14.json'
   cdblib.create (checkpoint_file)
   checkpoint = []
\end{cadabra}
\egroup

\clearpage

% ============================================================================================
\section*{Example 14 The Weyl tensor is conformally invariant}

This example shows that the Weyl tensor is conformally invariant. That is, for a pair of
metrics $g$ and $\overline{g}$ related by a conformal transformation,
$\overline{g}_{a b} = \phi g_{a b}$ then $\overline{C}^{a}_{b c d} = C^{a}_{b c d}$
or equally $\overline{C}_{a b c d} = \phi C_{a b c d}$.

\begin{cadabra}
   {a,b,c,d,e,f,g,h,i,j,k,l,m,n,o,p,q,r,s,u,v,w#}::Indices (position=independent).

   \partial{#}::PartialDerivative.

   g_{a b}::Metric.
   g^{a b}::InverseMetric.
   g_{a}^{b}::KroneckerDelta.

   GammaU := \Gamma^{a}_{b c} ->  1/2 g^{a d} (   \partial_{b}{g_{d c}}
                                                + \partial_{c}{g_{b d}}
                                                - \partial_{d}{g_{b c}}).

   GammaD := \Gamma_{a b c} ->  1/2 (   \partial_{b}{g_{a c}}
                                      + \partial_{c}{g_{b a}}
                                      - \partial_{a}{g_{b c}}).

   Rabcd := R_{a b c d} ->   \partial_{c}{\Gamma_{a b d}}
                           - \partial_{d}{\Gamma_{a b c}}
                           + \Gamma_{e a d} \Gamma^{e}_{b c}
                           - \Gamma_{e a c} \Gamma^{e}_{b d}.

   Rab     := R_{a b} -> g^{c d} R_{a c b d}.

   Rscalar := R -> g^{a b} R_{a b}.

   # Weyl in 4-dimensions

   Cabcd := R_{a b c d} - (1/2) (R_{a c} g_{b d} - R_{a d} g_{b c})
                        - (1/2) (g_{a c} R_{b d} - g_{a d} R_{b c})
                        + (R/6) (g_{a c} g_{b d} - g_{a d} g_{b c}).

   {\partial_{a b}{\phi},\partial_{a}{\phi},\phi}::SortOrder.
   {\partial_{a b}{g_{c d}},\partial_{a}{g_{b c}},g_{a b},g^{a b}}::SortOrder.

   substitute (Cabcd,Rscalar)
   substitute (Cabcd,Rab)
   substitute (Cabcd,Rabcd)
   substitute (Cabcd,GammaU)
   substitute (Cabcd,GammaD)

   distribute     (Cabcd)

   sort_product   (Cabcd)
   rename_dummies (Cabcd)
   canonicalise   (Cabcd)

   # this is the Weyl tensor on the base metric
   baseC := @(Cabcd).

   conformal := {g_{a b} -> \phi g_{a b}, g^{a b} -> (1/phi) g^{a b}}.

   substitute     (Cabcd, conformal)
   product_rule   (Cabcd)
   distribute     (Cabcd)
   product_rule   (Cabcd)
   distribute     (Cabcd)

   map_sympy      (Cabcd, "simplify")

   sort_product   (Cabcd)
   rename_dummies (Cabcd)
   canonicalise   (Cabcd)

   # this is the Weyl tensor on the conformal metric
   confC := @(Cabcd).

   # their difference, should be zero
   diff := @(confC) - \phi @(baseC).  # cdb (ex-14.diff.100,diff)

   distribute     (diff)
   sort_product   (diff)
   rename_dummies (diff)
   canonicalise   (diff)  # cdb (ex-14.diff.101,diff)

   # this trick is not essential but it does reduce the number of terms in diff
   substitute     (diff,$\partial_{a}{\partial_{b}{g_{c d}}} -> g_{c d b a}$)
   substitute     (diff,$\partial_{a}{g_{b c}} -> 0$)
   substitute     (diff,$g_{c d b a} -> \partial_{a}{\partial_{b}{g_{c d}}}$)  # cdb (ex-14.diff.102,diff)

   # standard expressions in 4-d
   substitute     (diff,$g_{a b} g^{a b} -> 4$,repeat=True)          # cdb (ex-14.diff.201,diff)
   substitute     (diff,$g_{a b} g^{c b} -> g_{a}^{c}$,repeat=True)  # cdb (ex-14.diff.202,diff)
   substitute     (diff,$g_{b a} g^{b c} -> g_{a}^{c}$,repeat=True)  # cdb (ex-14.diff.203,diff)
   substitute     (diff,$g_{a}^{a} -> 4$,repeat=True)                # cdb (ex-14.diff.204,diff)
   substitute     (diff,$g^{a}_{a} -> 4$,repeat=True)                # cdb (ex-14.diff.205,diff)
   eliminate_kronecker (diff)                                        # cdb (ex-14.diff.206,diff)

   # need a second round since the above block introduces new terms that match those just eliminated
   substitute     (diff,$g_{a b} g^{a b} -> 4$,repeat=True)          # cdb (ex-14.diff.301,diff)
   substitute     (diff,$g_{a b} g^{c b} -> g_{a}^{c}$,repeat=True)  # cdb (ex-14.diff.302,diff)
   substitute     (diff,$g_{b a} g^{b c} -> g_{a}^{c}$,repeat=True)  # cdb (ex-14.diff.303,diff)
   substitute     (diff,$g_{a}^{a} -> 4$,repeat=True)                # cdb (ex-14.diff.304,diff)
   substitute     (diff,$g^{a}_{a} -> 4$,repeat=True)                # cdb (ex-14.diff.305,diff)
   eliminate_kronecker (diff)                                        # cdb (ex-14.diff.306,diff)

   sort_product   (diff)
   rename_dummies (diff)
   canonicalise   (diff)  # cdb (ex-14.diff.400,diff)

   checkpoint.append (baseC)
   checkpoint.append (confC)
\end{cadabra}

\clearpage

\begin{dgroup*}
   \Dmath*{ \Delta = \Cdb*{ex-14.diff.102}}
\end{dgroup*}

\begin{dgroup*}
   \Dmath*{ \Delta = \Cdb*{ex-14.diff.201}}
\end{dgroup*}

\begin{dgroup*}
   \Dmath*{ \Delta = \Cdb*{ex-14.diff.202}}
\end{dgroup*}

\begin{dgroup*}
   \Dmath*{ \Delta = \Cdb*{ex-14.diff.203}}
\end{dgroup*}

\begin{dgroup*}
   \Dmath*{ \Delta = \Cdb*{ex-14.diff.204}}
\end{dgroup*}

\begin{dgroup*}
   \Dmath*{ \Delta = \Cdb*{ex-14.diff.205}}
\end{dgroup*}

\begin{dgroup*}
   \Dmath*{ \Delta = \Cdb*{ex-14.diff.206}}
\end{dgroup*}

\begin{dgroup*}
   \Dmath*{ \Delta = \Cdb*{ex-14.diff.301}}
\end{dgroup*}

\begin{dgroup*}
   \Dmath*{ \Delta = \Cdb*{ex-14.diff.302}}
\end{dgroup*}

\begin{dgroup*}
   \Dmath*{ \Delta = \Cdb*{ex-14.diff.303}}
\end{dgroup*}

\begin{dgroup*}
   \Dmath*{ \Delta = \Cdb*{ex-14.diff.304}}
\end{dgroup*}

\begin{dgroup*}
   \Dmath*{ \Delta = \Cdb*{ex-14.diff.305}}
\end{dgroup*}

\begin{dgroup*}
   \Dmath*{ \Delta = \Cdb*[\hskip2.0cm\hfill]{ex-14.diff.306}}
\end{dgroup*}

\begin{dgroup*}
   \Dmath*{ \Delta = \Cdb*{ex-14.diff.400}}
\end{dgroup*}

\clearpage

% ============================================================================================
% export to json format

\bgroup
\CdbSetup{action=hide}
\begin{cadabra}
   for i in range( len(checkpoint) ):
      cdblib.put ('check{:03d}'.format(i),checkpoint[i],checkpoint_file)
\end{cadabra}
\egroup

\end{document}
