\def\Date{10 Aug 2022}

\documentclass[12pt]{cdblatex}
\usepackage{examples}

\begin{document}

% ============================================================================================
\bgroup
\CdbSetup{action=hide}
\begin{cadabra}
   import cdblib
   checkpoint_file = 'tests/semantic/output/example-15.json'
   cdblib.create (checkpoint_file)
   checkpoint = []
\end{cadabra}
\egroup

\clearpage

% ============================================================================================
\section*{Example 15 Verifying the BSSN equations}

This is short example verifies two of the main equations in the Phys Rev D paper
by Miguel Alcubierre, Bernd Brugmann etal. (Phys.Rev.D. (62) 044034 (2000)).

The code for the full set of BSSN equations can be found at
\url{https://github.com/leo-brewin/adm-bssn-equations}

\begin{cadabra}
   {a,b,c,d,e,f,i,j,k,l,m,n,o,p,q,r,s,u#}::Indices(position=independent,values={t,x,y,z}).
   {t,x,y,z}::Coordinate.

   \partial{#}::PartialDerivative.
   D{#}::Derivative.
   DBar{#}::Derivative.

   N::Depends(t,x,y,z).

   g_{a b}::Symmetric.
   g^{a b}::Symmetric.
   g_{a}^{b}::KroneckerDelta.
   g^{a}_{b}::KroneckerDelta.

   g_{a b}::Depends(t,x,y,z).
   g^{a b}::Depends(t,x,y,z).

   gBar_{a b}::Symmetric.
   gBar^{a b}::Symmetric.
   gBar_{a}^{b}::KroneckerDelta.
   gBar^{a}_{b}::KroneckerDelta.

   gBar_{a b}::Depends(t,x,y,z).
   gBar^{a b}::Depends(t,x,y,z).

   trK::LaTeXForm("K").
   detg::LaTeXForm("g").
   ABar{#}::LaTeXForm("{\bar{A}}").
   DBar{#}::LaTeXForm("{\bar{D}}").
\end{cadabra}

\clearpage

% --------------------------------------------------------------------------------------------
\subsection*{15.1 Evolution equation for $\phi$}

\begin{cadabra}
   phi     := \phi -> (1/12) \log(detg).
   gdotK   := g^{i j} K_{i j} -> trK.
   DdetgDt := \partial_{t}{detg} -> detg g^{i j} \partial_{t}{g_{i j}}.

   DgijDt  := \partial_{t}{g_{i j}} -> -2 N K_{i j}.

   dlog    := \partial_{a?}{\log(A?)} -> (1/A?)\partial_{a?}{A?}.
   dexp    := \partial_{a?}{\exp(A?)} -> \exp(A?)\partial_{a?}{A?}.

   dotphi  := \partial_{t}{\phi}.

   substitute (dotphi, phi)                 # cdb (ex-15-02.101,dotphi)
   substitute (dotphi, dlog)                # cdb (ex-15-02.102,dotphi)
   substitute (dotphi, DdetgDt)             # cdb (ex-15-02.103,dotphi)
   substitute (dotphi, DgijDt)              # cdb (ex-15-02.104,dotphi)
   substitute (dotphi, gdotK)               # cdb (ex-15-02.105,dotphi)
   map_sympy  (dotphi, "simplify")          # cdb (ex-15-02.106,dotphi)

   DphiDt := \partial_{t}{\phi} -> @(dotphi).

   checkpoint.append (dotphi)
\end{cadabra}

\begin{align*}
  \frac{d{\phi}}{dt} &=\Cdb{ex-15-02.101}\\[10pt]
                     &=\Cdb{ex-15-02.102}\\[10pt]
                     &=\Cdb{ex-15-02.103}\\[10pt]
                     &=\Cdb{ex-15-02.104}\\[10pt]
                     &=\Cdb{ex-15-02.105}\\[10pt]
                     &=\Cdb{ex-15-02.106}
\end{align*}

\clearpage

% --------------------------------------------------------------------------------------------
\subsection*{15.2 Evolution equation for ${\bar g}_{ij}$}

\begin{cadabra}
   gBarij := gBar_{i j} -> \exp(-4\phi) g_{i j}.
   Kij    := K_{i j} -> A_{i j} + (1/3) g_{i j} trK.
   A2ABar := \exp(-4\phi) A_{i j} -> ABar_{i j}.
   ABar2A := ABar_{i j} -> \exp(-4\phi) A_{i j}.

   dotgBarij := \partial_{t}{gBar_{i j}}.

   substitute   (dotgBarij, gBarij)         # cdb (ex-15-03.101,dotgBarij)
   product_rule (dotgBarij)                 # cdb (ex-15-03.102,dotgBarij)
   substitute   (dotgBarij, dexp)           # cdb (ex-15-03.103,dotgBarij)
   substitute   (dotgBarij, DgijDt)         # cdb (ex-15-03.104,dotgBarij)
   substitute   (dotgBarij, DphiDt)         # cdb (ex-15-03.105,dotgBarij)
   substitute   (dotgBarij, Kij)            # cdb (ex-15-03.106,dotgBarij)
   distribute   (dotgBarij)                 # cdb (ex-15-03.107,dotgBarij)
   map_sympy    (dotgBarij, "simplify")     # cdb (ex-15-03.108,dotgBarij)
   substitute   (dotgBarij, A2ABar)         # cdb (ex-15-03.109,dotgBarij)

   DgBarijDt := \partial_{t}{gBar_{i j}} -> @(dotgBarij).

   checkpoint.append (dotgBarij)
\end{cadabra}

\begin{align*}
  \frac{d{\bar g}_{ij}}{dt} &=\Cdb{ex-15-03.101}\\[10pt]
                            &=\Cdb{ex-15-03.102}\\[10pt]
                            &=\Cdb{ex-15-03.103}\\[10pt]
                            &=\Cdb{ex-15-03.104}\\[10pt]
                            &=\Cdb{ex-15-03.105}\\[10pt]
                            &=\Cdb{ex-15-03.106}\\[10pt]
                            &=\Cdb{ex-15-03.107}\\[10pt]
                            &=\Cdb{ex-15-03.108}\\[10pt]
                            &=\Cdb{ex-15-03.109}
\end{align*}

\clearpage

% ============================================================================================
% export to json format

\bgroup
\CdbSetup{action=hide}
\begin{cadabra}
   for i in range( len(checkpoint) ):
      cdblib.put ('check{:03d}'.format(i),checkpoint[i],checkpoint_file)
\end{cadabra}
\egroup

\end{document}
