\documentclass[12pt]{cdblatex}
\usepackage{examples}
\usepackage{fancyhdr}
\usepackage{footer}

\begin{document}

% ============================================================================================
\bgroup
\CdbSetup{action=hide}
\begin{cadabra}
   import cdblib
   checkpoint_file = 'tests/semantic/output/example-03.json'
   cdblib.create (checkpoint_file)
   checkpoint = []
\end{cadabra}
\egroup

\clearpage

% ============================================================================================
\section*{Example 3a The Riemann curvature tensor}

\begin{cadabra}
   {a,b,c,d,e,f,g,h,i,j,k,l,m,n,o,p,q,r,s,t,u#}::Indices(position=independent).

   \partial{#}::PartialDerivative.

   \Gamma^{a}_{b c}::TableauSymmetry(shape={2}, indices={1,2});

   ;::Symbol;  # Suggsted by Kasper as a way to make use of ; legal
               # see https://cadabra.science/qa/473/is-this-legal-syntax
               # this code works with and without this trick

   # rules for the first two covariant derivs of V^a

   deriv1 := V^{a}_{; b}      -> \partial_{b}{V^{a}}
                               + \Gamma^{a}_{c b} V^{c}.        # cdb (ex-03.101,deriv1)

   deriv2 := V^{a}_{; b ; c}  -> \partial_{c}{V^{a}_{; b}}
                               + \Gamma^{a}_{d c} V^{d}_{; b}
                               - \Gamma^{d}_{b c} V^{a}_{; d}.  # cdb (ex-03.102,deriv2)

   substitute (deriv2,deriv1)                   # cdb (ex-03.103, deriv2)

   Vabc := V^{a}_{; b ; c} - V^{a}_{; c ; b}.   # cdb (ex-03.104, Vabc)

   substitute (Vabc,deriv2)                     # cdb (ex-03.105, Vabc)

   distribute     (Vabc)                        # cdb (ex-03.106, Vabc)
   product_rule   (Vabc)                        # cdb (ex-03.107, Vabc)

   sort_product   (Vabc)                        # cdb (ex-03.108, Vabc)
   rename_dummies (Vabc)                        # cdb (ex-03.109, Vabc)
   canonicalise   (Vabc)                        # cdb (ex-03.110, Vabc)

   sort_sum       (Vabc)                        # cdb (ex-03.111, Vabc)
   factor_out     (Vabc,$V^{a?}$)               # cdb (ex-03.112, Vabc)

   checkpoint.append (Vabc)

   # create rule for Riemann, export later (for use by lib/dgeom)

   substitute (Vabc,$V^{a} -> -1$)              # cdb (ex-03.113, Vabc)
                                                # note use of -1 to get correct
                                                # signs when coupled with the rule
                                                # for Rabcd (next statement)

   Rabcd := R^{a}_{d b c} -> @(Vabc).           # cdb (ex-03.114, Rabcd) #

   foo   := R^{a}_{b c d}.                      # cdb (ex-03.115, foo)
   substitute (foo, Rabcd)                      # cdb (ex-03.116, foo)

   # update rule to use nice indices

   Rabcd := R^{a}_{b c d} -> @(foo).

   checkpoint.append (Rabcd)

\end{cadabra}

\begin{dgroup*}[spread={3pt}]
   \Dmath*{\Cdb*{ex-03.101}}
   \Dmath*{\Cdb*{ex-03.102}}
   \Dmath*{\Cdb*{ex-03.103}}
\end{dgroup*}
%
\begin{dgroup*}[spread={3pt}]
   \Dmath*{\cdb{ex-03.104} = \Cdb*[\V{5pt}\hskip2.0cm\hfill]{ex-03.105}
                           = \Cdb*[\V{5pt}\hskip2.0cm\hfill]{ex-03.106}
                           = \Cdb*[\V{5pt}\hskip2.5cm\hfill]{ex-03.107}
                           = \Cdb*[\V{5pt}\hskip2.5cm\hfill]{ex-03.108}
                           = \Cdb*[\V{5pt}\hskip2.5cm\hfill]{ex-03.109}
                           = \Cdb*[\V{5pt}\hfill]{ex-03.110}
                           = \Cdb*[\V{5pt}\hfill]{ex-03.111}
                           = \Cdb*[\V{5pt}\hfill]{ex-03.112}
                           = - R^{a}{}_{dbc} V^d\quad\text{(MTW)}}
\end{dgroup*}

\Dmath*{\cdb{ex-03.115} = \cdb{ex-03.116}}

\clearpage

% ============================================================================================
\section*{Example 3b The Riemann curvature tensor}

This differs from the above by not using the \verb|::TableauSymmetry| property.
It gives the same results as above but it does require a little bit more
housekeeping.

\begin{cadabra}
   {a,b,c,d,e,f,g,h,i,j,k,l,m,n,o,p,q,r,s,t,u#}::Indices(position=independent).

   \partial{#}::PartialDerivative.

   ;::Symbol;  # Suggsted by Kasper as a way to make use of ; legal
               # see https://cadabra.science/qa/473/is-this-legal-syntax
               # this code works with and without this trick

   # rules for the first two covariant derivs of V^a

   deriv1 := V^{a}_{; b}      -> \partial_{b}{V^{a}}
                               + \Gamma^{a}_{c b} V^{c}.        # cdb (ex-03.301,deriv1)

   deriv2 := V^{a}_{; b ; c}  -> \partial_{c}{V^{a}_{; b}}
                               + \Gamma^{a}_{d c} V^{d}_{; b}
                               - \Gamma^{d}_{b c} V^{a}_{; d}.  # cdb (ex-03.302,deriv2)

   substitute (deriv2,deriv1)                    # cdb (ex-03.303, deriv2)

   Vabc := V^{a}_{; b ; c} - V^{a}_{; c ; b}.    # cdb (ex-03.304, Vabc)

   substitute (Vabc,deriv2)                      # cdb (ex-03.305, Vabc)

   distribute     (Vabc)                         # cdb (ex-03.306, Vabc)
   product_rule   (Vabc)                         # cdb (ex-03.307, Vabc)

   # ------------------------------------------------------
   # trick to obtain a symmetric connection

   G_{a b}::Symmetric.

   substitute     (Vabc,$\Gamma^{a}_{b c} ->  G^{a} G_{b c}$)
   sort_product   (Vabc)                         # cdb (ex-03.308, Vabc)
   rename_dummies (Vabc)                         # cdb (ex-03.309, Vabc)
   canonicalise   (Vabc)                         # cdb (ex-03.310, Vabc)
   substitute     (Vabc,$G^{a} G_{b c} -> \Gamma^{a}_{b c}$,repeat=True)
   # ------------------------------------------------------

   sort_product   (Vabc)
   rename_dummies (Vabc)
   canonicalise   (Vabc)

   sort_sum       (Vabc)                         # cdb (ex-03.311, Vabc)
   factor_out     (Vabc,$V^{a?}$)                # cdb (ex-03.312, Vabc)

   checkpoint.append (Vabc)
\end{cadabra}

\begin{dgroup*}[spread={3pt}]
   \Dmath*{\Cdb*{ex-03.301}}
   \Dmath*{\Cdb*{ex-03.302}}
   \Dmath*{\Cdb*{ex-03.303}}
\end{dgroup*}
%
\begin{dgroup*}[spread={3pt}]
   \Dmath*{\cdb{ex-03.304} = \Cdb*[\V{5pt}\hskip2.0cm\hfill]{ex-03.305}
                           = \Cdb*[\V{5pt}\hskip2.0cm\hfill]{ex-03.306}
                           = \Cdb*[\V{5pt}\hskip2.5cm\hfill]{ex-03.307}
                           = \Cdb*[\V{5pt}\hskip2.5cm\hfill]{ex-03.308}
                           = \Cdb*[\V{5pt}\hskip2.5cm\hfill]{ex-03.309}
                           = \Cdb*[\V{5pt}\hfill]{ex-03.310}
                           = \Cdb*[\V{5pt}\hfill]{ex-03.311}
                           = \Cdb*[\V{5pt}\hfill]{ex-03.312}
                           = - R^{a}{}_{dbc} V^d\quad\text{(MTW)}}
\end{dgroup*}

\clearpage

% ============================================================================================
% export to json format

\bgroup
\CdbSetup{action=hide}
\begin{cadabra}
   for i in range( len(checkpoint) ):
      cdblib.put ('check{:03d}'.format(i),checkpoint[i],checkpoint_file)
\end{cadabra}
\egroup

\end{document}
