\documentclass[12pt]{cdblatex}
\usepackage{examples}

\begin{document}

% ============================================================================================
\bgroup
\CdbSetup{action=hide}
\begin{cadabra}
   import cdblib
   checkpoint_file = 'tests/semantic/output/example-11.json'
   cdblib.create (checkpoint_file)
   checkpoint = []
\end{cadabra}
\egroup

\clearpage

% ============================================================================================
\section*{Example 11 The RNC connection.}

\begin{cadabra}
   {a,b,c,d,e,f,g,h,i,j,k,l,m,n,o,p,q,r,s,u#}::Indices(position=independent).

   D{#}::PartialDerivative.
   \nabla{#}::Derivative.

   g_{a b}::Metric.
   g^{a b}::InverseMetric.
   g^{a b}::Weight(label=gnum,value=1).

   \delta{#}::KroneckerDelta.

   R_{a b c d}::RiemannTensor.
   R_{a b c d}::Depends(\nabla{#}).

   x^{a}::Depends(D{#}).
   x^{a}::Weight(label=xnum,value=1).

   Dx := D_{a}{x^{b}} -> \delta^{b}_{a}.   # cdb (ex-11.000,Dx)

   gab := g_{a b} -> g_{a b}
                  -  (1/3)   x^{c} x^{d} R_{a c b d}
                  -  (1/6)   x^{c} x^{d} x^{e} \nabla_{c}{R_{a d b e}}
                  +  (1/180) x^{c} x^{d} x^{e} x^{f} ( 8 g^{g h} R_{a c d g} R_{b e f h}
                                                      -9 \nabla_{c d}{R_{a e b f}} ).      # cdb (ex-11.001,gab)

   iab := g^{a b} -> g^{a b}
                  + (1/3)  x^{c} x^{d} g^{a e} g^{b f} R_{c e d f}
                  + (1/6)  x^{c} x^{d} x^{e} g^{a f} g^{b g} \nabla_{c}{R_{d f e g}}
                  + (1/60) x^{c} x^{d} x^{e} x^{f} g^{a g} g^{b h}
                                         ( 4 g^{i j} R_{c g d i} R_{e h f j}
                                          +3 \nabla_{c d}{R_{e g f h}} ).                  # cdb(ex-11.002,iab)

   distribute (gab)
   distribute (iab)

   ChrSym := \Gamma^{a}_{b c} -> 1/2 g^{a d} (  D_{b}{g_{d c}}
                                              + D_{c}{g_{b d}}
                                              - D_{d}{g_{b c}} ).   # cdb (ex-11.003,ChrSym)

   Gamma := \Gamma^{a}_{b c}.              # cdb (ex-11.100,Gamma)

   substitute     (Gamma,ChrSym)           # cdb (ex-11.101,Gamma)
   substitute     (Gamma,gab)              # cdb (ex-11.102,Gamma)
   substitute     (Gamma,iab)              # cdb (ex-11.103,Gamma)
   distribute     (Gamma)                  # cdb (ex-11.104,Gamma)
   unwrap         (Gamma)                  # cdb (ex-11.105,Gamma)
   product_rule   (Gamma)                  # cdb (ex-11.106,Gamma)
   distribute     (Gamma)                  # cdb (ex-11.107,Gamma)
   substitute     (Gamma,Dx)               # cdb (ex-11.108,Gamma)
   eliminate_kronecker (Gamma)             # cdb (ex-11.109,Gamma)

   def truncate (obj,n):

       ans = Ex("0")  # create a Cadabra object with value zero

       for i in range (0,n+1):
          foo := @(obj).
          bah  = Ex("xnum = " + str(i))
          distribute  (foo)
          keep_weight (foo, bah)
          ans = ans + foo

       return ans

   checkpoint.append (Gamma)

   # sort_product   (Gamma)  # 52.3 sec, 49 Mbyte
   # rename_dummies (Gamma)  # 58.6 sec, 51 Mbyte
   # canonicalise   (Gamma)  # killed after 20 mins and over 500 Mbyte

   Gamma = truncate (Gamma,3)   # cdb (ex-11.110,Gamma)  # allow up to 3rd order in x^a

   sort_product   (Gamma)
   rename_dummies (Gamma)
   canonicalise   (Gamma)

   checkpoint.append (Gamma)

   # ==========================================================================
   # the remaining code is just for pretty printing

   def product_sort (obj):
       substitute (obj,$ g^{a b}                   -> A001^{a b}                $)
       substitute (obj,$ x^{a}                     -> A002^{a}                  $)
       substitute (obj,$ z^{a}                     -> A003^{a}                  $)
       substitute (obj,$ R_{a b c d}               -> A004_{a b c d}            $)
       substitute (obj,$ \nabla_{e}{R_{a b c d}}   -> A005_{a b c d e}          $)
       substitute (obj,$ \nabla_{e f}{R_{a b c d}} -> A006_{a b c d e f}        $)
       sort_sum       (obj)
       sort_product   (obj)
       rename_dummies (obj)
       substitute (obj,$ A001^{a b}                -> g^{a b}                   $)
       substitute (obj,$ A002^{a}                  -> x^{a}                     $)
       substitute (obj,$ A003^{a}                  -> z^{a}                     $)
       substitute (obj,$ A004_{a b c d}            -> R_{a b c d}               $)
       substitute (obj,$ A005_{a b c d e}          -> \nabla_{e}{R_{a b c d}}   $)
       substitute (obj,$ A006_{a b c d e f}        -> \nabla_{e f}{R_{a b c d}} $)

   def get_xterm (obj,n):

       foo := @(obj).
       bah  = Ex("xnum = " + str(n))
       distribute  (foo)
       keep_weight (foo, bah)

       return foo

   def get_gterm (obj,n):

       foo := @(obj).
       bah  = Ex("gnum = " + str(n))
       distribute  (foo)
       keep_weight (foo, bah)

       return foo

   def reformat (obj,scale):

      foo  = Ex(str(scale))
      bah := @(foo) @(obj).

      distribute     (bah)
      product_sort   (bah)
      rename_dummies (bah)
      canonicalise   (bah)
      factor_out     (bah,$x^{a?},g^{b? c?}$)
      ans := @(bah) / @(foo).

      return ans

   gam1  = get_xterm (Gamma, 1)                          # cdb (ex-11.200,gam1)
   gam2  = get_xterm (Gamma, 2)                          # cdb (ex-11.201,gam2)
   gam3  = get_xterm (Gamma, 3)                          # cdb (ex-11.202,gam3)

   gam31 = get_gterm (gam3,  1)                          # cdb (ex-11.210,gam31)
   gam32 = get_gterm (gam3,  2)                          # cdb (ex-11.211,gam31)

   gam1  = reformat (gam1,   3)                          # cdb (ex-11.220,gam1)
   gam2  = reformat (gam2,  12)                          # cdb (ex-11.221,gam2)

   gam31 = reformat (gam31, 40)                          # cdb (ex-11.222,gam31)
   gam32 = reformat (gam32, 45)                          # cdb (ex-11.223,gam32)

   Gamma  := @(gam1) + @(gam2) + @(gam31) + @(gam32).    # cdb (ex-11.230,Gamma)
   Scaled := 360 @(Gamma).                               # cdb (ex-11.231,Scaled)

   checkpoint.append (Gamma)

\end{cadabra}

\clearpage

\begin{dgroup*}
   \Dmath*{\cdb{ex-11.100}(x) = \Cdb*[\hskip 3.0cm]{ex-11.230}}
\end{dgroup*}

\begin{dgroup*}
   \Dmath*{360 \cdb{ex-11.100}(x) = \Cdb*[\hskip 3.0cm]{ex-11.231}}
\end{dgroup*}

\clearpage

% ============================================================================================
% export to json format

% --------------------------------------------------------------------------------------------
\subsection*{Save $\Gamma^{a}{}_{bc}$ for later use in Example 12.}

\begin{cadabra}
   jsonfile = 'example-11.json'
   cdblib.create (jsonfile)
   cdblib.put ('Gamma',Gamma,jsonfile)
\end{cadabra}

\bgroup
\CdbSetup{action=hide}
\begin{cadabra}
   for i in range( len(checkpoint) ):
      cdblib.put ('check{:03d}'.format(i),checkpoint[i],checkpoint_file)
\end{cadabra}
\egroup

\end{document}
