\def\Date{10 Aug 2022}

\documentclass[12pt]{cdblatex}
\usepackage{examples}

\begin{document}

% ============================================================================================
\section*{Example 11 The RNC connection.}

\CdbSetup{action=hide}

\begin{cadabra}
   {a,b,c,d,e,f,g,h,i,j,k,l,m,n,o,p,q,r,s,u#}::Indices(position=independent).

   D{#}::PartialDerivative.
   \nabla{#}::Derivative.

   g_{a b}::Metric.
   g^{a b}::InverseMetric.

   \delta{#}::KroneckerDelta.

   R_{a b c d}::RiemannTensor.

   def truncate (obj,n):

       ans = Ex("0")

       for i in range (0,n+1):
          foo := @(obj).
          bah  = Ex("xnum = " + str(i))
          distribute  (foo)
          keep_weight (foo, bah)
          ans = ans + foo

       return ans

   def check (objA,objB):

       tmp := @(objA) - @(objB).

       distribute     (tmp)
       sort_product   (tmp)
       rename_dummies (tmp)
       canonicalise   (tmp)

       return tmp

   import cdblib

   # ---------------------------------------------
   check000A = truncate ( cdblib.get ('check000','expected/example-11.json'), 3)
   check001A = truncate ( cdblib.get ('check001','expected/example-11.json'), 3)
   check002A = truncate ( cdblib.get ('check002','expected/example-11.json'), 3)

   # ---------------------------------------------
   check000B = truncate ( cdblib.get ('check000','output/example-11.json'), 3)
   check001B = truncate ( cdblib.get ('check001','output/example-11.json'), 3)
   check002B = truncate ( cdblib.get ('check002','output/example-11.json'), 3)

   # ---------------------------------------------
   diff000 = check (check000A,check000B)   # cdb (ex-11.diff000,diff000)
   diff001 = check (check001A,check001B)   # cdb (ex-11.diff001,diff001)
   diff002 = check (check002A,check002B)   # cdb (ex-11.diff002,diff002)

\end{cadabra}

% \clearpage

\begin{dgroup*}
   \Dmath*{\cdb*{ex-11.diff000}}
   \Dmath*{\cdb*{ex-11.diff001}}
   \Dmath*{\cdb*{ex-11.diff002}}
\end{dgroup*}

\end{document}
