\documentclass[12pt]{cdblatex}
\usepackage{fancyhdr}
\usepackage{footer}

\begin{document}

% ============================================================================================
\section*{Example 12 Checking the 2nd and 3rd order terms of Calzetta etal.}

\CdbSetup{action=hide}

\begin{cadabra}
   {a,b,c,d,e,f,g,h,i,j,k,l,m,n,o,p,u,v#}::Indices("latin",position=independent).
   {\mu,\nu,\rho,\sigma,\tau,\lambda,\xi#}::Indices("greek",position=independent).

   \nabla{#}::Derivative.

   g_{a b}::Metric.
   g^{a b}::InverseMetric.

   \delta{#}::KroneckerDelta.

   R_{a b c d}::RiemannTensor.

   def check (objA,objB):

       tmp := @(objA) - @(objB).

       distribute     (tmp)
       sort_product   (tmp)
       rename_dummies (tmp)
       canonicalise   (tmp)

       return tmp

   import cdblib

   # ---------------------------------------------
   check000A = cdblib.get ('check000','expected/example-12.json')
   check001A = cdblib.get ('check001','expected/example-12.json')
   check002A = cdblib.get ('check002','expected/example-12.json')
   check003A = cdblib.get ('check003','expected/example-12.json')

   # ---------------------------------------------
   check000B = cdblib.get ('check000','output/example-12.json')
   check001B = cdblib.get ('check001','output/example-12.json')
   check002B = cdblib.get ('check002','output/example-12.json')
   check003B = cdblib.get ('check003','output/example-12.json')

   # ---------------------------------------------
   diff000 = check (check000A,check000B)   # cdb (ex-12.diff000,diff000)
   diff001 = check (check001A,check001B)   # cdb (ex-12.diff001,diff001)
   diff002 = check (check002A,check002B)   # cdb (ex-12.diff002,diff002)
   diff003 = check (check003A,check003B)   # cdb (ex-12.diff003,diff003)

\end{cadabra}

% \clearpage

\begin{dgroup*}
   \Dmath*{\cdb*{ex-12.diff000}}
   \Dmath*{\cdb*{ex-12.diff001}}
   \Dmath*{\cdb*{ex-12.diff002}}
   \Dmath*{\cdb*{ex-12.diff002}}
\end{dgroup*}

\end{document}
